COFFEE: an Optimizing Compiler\\for Finite Element Local Assembly

### 2014-04-14 paper/bare_conf.tex
Update licm table
Clean up

\begin{algorithm}[t]
\footnotesize
\KwSty{void} burgers(double A[12][12], double **c, double **w) $\lbrace$\\
~~// K, det = Compute Jacobian (c) \\
~~\\
~~\KwSty{static const double} W5[5] = $\lbrace$...$\rbrace$\\
~~\KwSty{static const double} X1$\_$D001[5][12] = $\lbrace\lbrace$...$\rbrace\rbrace$\\
~~\KwSty{static const double} X2$\_$D001[5][12] = $\lbrace\lbrace$...$\rbrace\rbrace$\\
~~//11 other basis functions definitions.\\
~~...\\
~~\\
~~\KwSty{for} (\KwSty{int} i = 0; i$<$5; i++) $\lbrace$\\
~~~~\KwSty{double} F0 = 0.0;\\
~~~~//10 other declarations (F1, F2,...)\\
~~~~...\\
~~~~\KwSty{for} (\KwSty{int} r = 0; r$<$12; r++) $\lbrace$\\
~~~~~~F0 += (w[r][0]*X1$\_$D100[i][r]);\\
~~~~~~//10 analogous statements (F1, F2, ...)\\
~~~~$\rbrace$\\
~~~~...\\
~~~~\KwSty{for} (\KwSty{int} j = 0; j$<$12; j++) \\
~~~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$12; k++) \\
~~~~~~~~A[j][k] += (..(K5*F9)+(K8*F10))*Y1[i][j])+\\
~~~~~~~~~+(((K0*X1$\_$D100[i][k])+(K3*X1$\_$D010[i][k])+\\
~~~~~~~~~+(K6*X1$\_$D001[i][k]))*Y2[i][j]))*F11)+\\
~~~~~~~~~+(..((K2*X2$\_$D100[i][k])+...+(K8*X2$\_$D001[i][k])*\\
~~~~~~~~~*(K2*X2$\_$D100[i][j])+...+(K8*X2$\_$D001[i][j]))..)+\\
~~~~~~~~~+ $<$roughly a hundred sum/muls go here$>$)..)*\\
~~~~~~~~~*det*W5[i]);\\
~~$\rbrace$ \\
$\rbrace$
\caption{Local assembly code generated by Firedrake for a Burgers problem on a 3D tetrahedral mesh using Lagrange $p=1$ elements.}
\label{code:burgers}
\end{algorithm}

@@

Experiments were run on two Intel architectures, a Sandy Bridge (I7-2600 CPU, running at 3.4GHz, 32KB L1 cache and 256KB L2 cache) and a Xeon Phi (5110P, running at 1.05Ghz in native mode, 32KB L1 cache and 512KB L2 cache). We have chosen these two architectures because of the differences in the number of logical registers and SIMD lanes, which can impact the effectiveness of the optimization strategy. All experiments are sequential. The \texttt{icc 13.1} compiler was used. On the Sandy Bridge, the compilation flags used were \texttt{-O2} and \texttt{-xAVX} for auto-vectorization. On the Xeon Phi, optimization level \texttt{-O3} was used. Other optimization levels performed, in general, slightly worse. Our code transformations were evaluated in three real-world problems based on the following PDEs:
\begin{itemize}
\item Helmholtz
\item Advection-Diffusion
\item Burgers
\end{itemize}
The code was written in UFL (available at~\cite{ufl-code}) and then executed over real unstructured meshes through Firedrake. The Helmholtz code has already been shown in Figure~\ref{code:helmholtz}. For Advection-Diffusion, the Diffusion equation, which uses the same differential operators as Helmholtz, is considered. In the Diffusion kernel, the main differences with respect to Helmholtz are the absence of the $Y$ array and the presence of a few more constants for computing the element matrix $A$. Burgers is a non-linear problem employing differential operators different from those of Helmholtz, which has a major impact on the generated assembly code (see Figure~\ref{code:burgers}), where a larger number of basis function matrices ($X1$, $X2$, ...) and constants ($F0$, $F1$, ..., $K0$, $K1$, ...) are needed.

@@

\subsection{Impact of Generalized Loop-invariant Code Motion}
%\newcommand{\massmatrixtwodresultsnorms}{
%\begin{tabularx}{1.0\textwidth}{p{3.5cm}p{3.5cm}|p{2.25cm}p{2.25cm}p{2.25cm}p{2.25cm}|}
%\cline{3-6}
%& & \multicolumn{2}{c|}{\texttt{Sandy Bridge}} & \multicolumn{2}{c|}{\texttt{Xeon Phi}}\\
%\cline{3-6}
%& & \multicolumn{1}{c}{\texttt{licm}} & \multicolumn{1}{c|}{\texttt{licm-ap}} & \multicolumn{1}{c}{\texttt{licm}} & \multicolumn{1}{c|}{\texttt{licm-ap}} \\
%\cline{1-6}
%%\multicolumn{1}{|c|}{\texttt{Helmholtz}} & \multicolumn{1}{c|}{\texttt{triangle}} & ~~1.34$\times$-4.32$\times$ & ~1.34$\times$-4.32$\times$ & ~~1.34$\times$-4.32$\times$ & ~1.34$\times$-4.32$\times$ \\
%\multicolumn{1}{|c|}{\texttt{Helmholtz}} & \multicolumn{1}{c|}{\texttt{triangle}} & ~~1.05$\times$-1.68$\times$ & \multicolumn{1}{c|}{~1.32$\times$-4.14$\times$} & ~~1.15$\times$-1.70$\times$ & ~~1.54$\times$-3.27$\times$ \\
%\multicolumn{1}{|c|}{\texttt{Helmholtz}} & \multicolumn{1}{c|}{\texttt{tetrahedron}} & ~~1.36$\times$-2.64$\times$ & \multicolumn{1}{c|}{~1.35$\times$-3.32$\times$} & ~~1.18$\times$-1.87$\times$ & ~~1.27$\times$-2.92$\times$ \\
%\multicolumn{1}{|c|}{\texttt{Helmholtz}} & \multicolumn{1}{c|}{\texttt{prism}} & ~~2.16$\times$-2.45$\times$ & \multicolumn{1}{c|}{~2.42$\times$-2.63$\times$} & ~~0.96$\times$-1.64$\times$ & ~~1.84$\times$-1.72$\times$ \\
%\cline{1-6}
%\multicolumn{1}{|c|}{\texttt{Diffusion}} & \multicolumn{1}{c|}{\texttt{triangle}} & ~~1.09$\times$-1.64$\times$ & \multicolumn{1}{c|}{~1.37$\times$-4.28$\times$} & ~~1.13$\times$-1.22$\times$ & ~~0.92$\times$-2.79$\times$ \\
%\multicolumn{1}{|c|}{\texttt{Diffusion}} & \multicolumn{1}{c|}{\texttt{tetrahedron}} & ~~1.30$\times$-3.11$\times$ & \multicolumn{1}{c|}{~1.41$\times$-3.82$\times$} & ~~1.08$\times$-2.11$\times$ & ~~1.10$\times$-5.53$\times$ \\
%\multicolumn{1}{|c|}{\texttt{Diffusion}} & \multicolumn{1}{c|}{\texttt{prism}} & ~~1.81$\times$-2.70$\times$ & \multicolumn{1}{c|}{~2.55$\times$-3.13$\times$} & ~~1.04$\times$-2.74$\times$ & ~~1.73$\times$-4.14$\times$ \\
%\cline{1-6}
%\multicolumn{1}{|c|}{\texttt{Burgers}} & \multicolumn{1}{c|}{\texttt{triangle}} & ~~1.53$\times$-2.46$\times$ & \multicolumn{1}{c|}{~1.56$\times$-2.77$\times$} & ~~1.19$\times$-2.97$\times$ & ~~2.70$\times$-4.27$\times$ \\
%\multicolumn{1}{|c|}{\texttt{Burgers}} & \multicolumn{1}{c|}{\texttt{tetrahedron}} & ~~1.58$\times$-2.24$\times$ & \multicolumn{1}{c|}{~1.59$\times$-2.10$\times$} & ~~1.03$\times$-1.21$\times$ & ~~1.31$\times$-1.55$\times$ \\
%\multicolumn{1}{|c|}{\texttt{Burgers}} & \multicolumn{1}{c|}{\texttt{prism}} & ~~1.32$\times$-2.11$\times$ & \multicolumn{1}{c|}{~1.42$\times$-2.31$\times$} & ~~1.04$\times$-1.40$\times$ & ~~1.27$\times$-2.18$\times$ \\
%\cline{1-6}
%\end{tabularx}
%}

\newcommand{\massmatrixtwodresultsnorms}{
\begin{tabularx}{1.0\textwidth}{cccccc|cccc|cccc|cccc}
\cline{3-18}
& & \multicolumn{8}{c}{\texttt{Sandy Bridge}} & \multicolumn{8}{c}{\texttt{Xeon Phi}}\\
\cline{3-18}
& & \multicolumn{4}{c}{\texttt{licm}} & \multicolumn{4}{c}{\texttt{licm-ap}} & \multicolumn{4}{c}{\texttt{licm}} & \multicolumn{4}{c}{\texttt{licm-ap}} \\
\cline{1-18}
& & \texttt{p1} & \texttt{p2} & \texttt{p3} & \texttt{p4} & \texttt{p1} & \texttt{p2} & \texttt{p3} & \texttt{p4} & \texttt{p1} & \texttt{p2} & \texttt{p3} & \texttt{p4} & \texttt{p1} & \texttt{p2} & \texttt{p3} & \texttt{p4} \\[0.1cm]
\texttt{Helmholtz} & \texttt{triangle} & 1.05 & 1.46 & 1.68 & 1.67 & 1.32 & 1.88 & 2.87 & 4.13 & 1.05 & 1.46 & 1.68 & 1.67 & 1.32 & 1.88 & 2.87 & 4.13 \\
\texttt{Helmholtz} & \texttt{tetrahedron} & 1.36 & 2.10 & 2.64 & 2.27 & 1.35 & 3.32 & 2.66 & 3.27 & 1.36 & 2.10 & 2.64 & 2.27 & 1.35 & 3.32 & 2.66 & 3.27 \\
\texttt{Helmholtz} & \texttt{prism} & 2.16 & 2.28 & 2.45 & 2.06 & 2.63 & 2.74 & 2.43 & 2.75 & 2.16 & 2.28 & 2.45 & 2.06 & 2.63 & 2.74 & 2.43 & 2.75\\[0.1cm]
%\cline{1-18}
\texttt{Diffusion} & \texttt{triangle} & 1.09 & 1.68 & 1.97 & 1.64 & 1.38 & 1.99 & 3.07 & 4.28 & 2.15 & 1.82 & 2.71 & 2.32 & 2.55 & 3.13 & 2.73 & 2.69\\
\texttt{Diffusion} & \texttt{tetrahedron} & 1.30 & 2.20 & 3.12 & 2.60 & 1.41 & 3.70 & 3.18 & 3.82 & 2.15 & 1.82 & 2.71 & 2.32 & 2.55 & 3.13 & 2.73 & 2.69\\
\texttt{Diffusion} & \texttt{prism} & 2.15 & 1.82 & 2.71 & 2.32 & 2.55 & 3.13 & 2.73 & 2.69 & 2.15 & 1.82 & 2.71 & 2.32 & 2.55 & 3.13 & 2.73 & 2.69\\[0.1cm]
%\cline{1-18}
\texttt{Burgers} & \texttt{triangle} & 1.53 & 1.81 & 2.68 & 2.46 & 1.56 & 2.28 & 2.61 & 2.77 & 1.53 & 1.81 & 2.68 & 2.46 & 1.56 & 2.28 & 2.61 & 2.77 \\
\texttt{Burgers} & \texttt{tetrahedron} & 1.61 & 2.24 & 1.69 & 1.59 & 1.61 & 2.10 & 1.60 & 1.78 & 1.53 & 1.81 & 2.68 & 2.46 & 1.56 & 2.28 & 2.61 & 2.77 \\
\texttt{Burgers} & \texttt{prism} & 2.11 & 2.20 & 1.66 & 1.32 & 2.19 & 2.32 & 1.64 & 1.42 & 1.53 & 1.81 & 2.68 & 2.46 & 1.56 & 2.28 & 2.61 & 2.77 \\
\cline{1-18}
\end{tabularx}
}

@@

Table~\ref{table:perf-results-licm} illustrates the performance improvement obtained on the Sandy Bridge and Xeon Phi machines when \emph{licm}, data alignment, and padding are used. We distinguish between \emph{licm} and \emph{licm-ap}; the latter makes use of padding and data alignment. Inspection of assembly code generated by \texttt{icc} confirmed all limitations described in Section~\ref{sec:code-transf-licm}: only sub-expressions invariant to the inner loop are hoisted and their execution is not vectorized. Padding and data alignment enhance, in general, the quality of auto-vectorization. Sometimes the run-time of \emph{licm-ap} is similar to that of \emph{licm} because the element matrix size is already a multiple of the vector length, so no scalar reminder loop has to be introduced. Occasionally \emph{licm-ap} is slower than \emph{licm} (e.g. in Burgers 3D p3 on the Sandy Bridge). One possible explanation is that the large number of aligned temporaries introduced by \emph{licm} induce cache associativity conflicts.

@@

\section{Related Work}
\label{sec:related-work}
The finite element method is extensively used to approximate solutions of PDEs. Well-known frameworks and applications include nek5000~\cite{nek5000-web-page}, the FEniCS project~\cite{Fenics}, Fluidity~\cite{fluidity_manual_v4}, and of course Firedrake. Numerical integration is usually employed to implement the local assembly phase. The recent introduction of DSLs to decouple the finite element specification from its underlying implementation facilitated, however, the development of novel approaches. Methods based on tensor contraction~\cite{FFC-TC} and symbolic manipulation~\cite{Francis} have been developed. We have designed COFFEE to specifically target numerical integration because it has been demonstrated that it remains the optimal choice for a wide class of problems~\cite{quadrature1}.

>>>>

\begin{algorithm}[t]
\footnotesize
\KwSty{void} burgers(double A[12][12], double **coords, double **w) $\lbrace$\\
~~// K, det = Compute Jacobian (coords) \\
~~\\
~~\KwSty{static const double} W5[5] = $\lbrace$...$\rbrace$\\
~~\KwSty{static const double} X1$\_$D001[5][12] = $\lbrace\lbrace$...$\rbrace\rbrace$\\
~~\KwSty{static const double} X2$\_$D001[5][12] = $\lbrace\lbrace$...$\rbrace\rbrace$\\
~~//11 other basis functions definitions.\\
~~...\\
~~\\
~~\KwSty{for} (\KwSty{int} i = 0; i$<$5; i++) $\lbrace$\\
~~~~\KwSty{double} F0 = 0.0;\\
~~~~//10 other declarations (F1, F2,...)\\
~~~~...\\
~~~~\KwSty{for} (\KwSty{int} r = 0; r$<$12; r++) $\lbrace$\\
~~~~~~F0 += (w[r][0]*X1$\_$D100[i][r]);\\
~~~~~~//10 analogous statements (F1, F2, ...)\\
~~~~$\rbrace$\\
~~~~...\\
~~~~\KwSty{for} (\KwSty{int} j = 0; j$<$12; j++) \\
~~~~~~\KwSty{for} (\KwSty{int} k = 0; k$<$12; k++) \\
~~~~~~~~A[j][k] += (..(K5*F9)+(K8*F10))*Y1[i][j])+\\
~~~~~~~~~+(((K0*X1$\_$D100[i][k])+(K3*X1$\_$D010[i][k])+\\
~~~~~~~~~+(K6*X1$\_$D001[i][k]))*Y2[i][j]))*F11)+\\
~~~~~~~~~+(..((K2*X2$\_$D100[i][k])+...+(K8*X2$\_$D001[i][k])*\\
~~~~~~~~~*(K2*X2$\_$D100[i][j])+...+(K8*X2$\_$D001[i][j]))..)+\\
~~~~~~~~~+ $<$roughly a hundred sum/muls go here$>$)..)*\\
~~~~~~~~~*det*W5[i]);\\
~~$\rbrace$ \\
$\rbrace$
\caption{Local assembly code generated by Firedrake for a Burgers problem on a 3D tetrahedral mesh using Lagrange $p=1$ elements.}
\label{code:burgers}
\end{algorithm}

@@

Experiments were run on a single core of two Intel architectures, a Sandy Bridge (I7-2600 CPU, running at 3.4GHz, 32KB L1 cache and 256KB L2 cache) and a Xeon Phi (5110P, running at 1.05Ghz in native mode, 32KB L1 cache and 512KB L2 cache). We have chosen these two architectures because of the differences in the number of logical registers and SIMD lanes, which can impact the effectiveness of the optimization strategy. The \texttt{icc 13.1} compiler was used. On the Sandy Bridge, the compilation flags used were \texttt{-O2} and \texttt{-xAVX} for auto-vectorization. On the Xeon Phi, optimization level \texttt{-O3} was used. Other optimization levels performed, in general, slightly worse. Our code transformations were evaluated in three real-world problems based on the following PDEs:
\begin{itemize}
\item Helmholtz
\item Advection-Diffusion
\item Burgers
\end{itemize}
The code was written in UFL (available at~\cite{ufl-code}) and then executed over real unstructured meshes through Firedrake. The Helmholtz code has already been shown in Figure~\ref{code:helmholtz}. For Advection-Diffusion, the Diffusion equation, which uses the same differential operators as Helmholtz, is considered. In the Diffusion kernel, the main differences with respect to Helmholtz are the absence of the $Y$ array and the presence of a few more constants for computing the element matrix $A$. Burgers is a non-linear problem employing differential operators different from those of Helmholtz, which has a major impact on the generated assembly code (see Figure~\ref{code:burgers}), where a larger number of basis function matrices ($X1$, $X2$, ...) and constants ($F0$, $F1$, ..., $K0$, $K1$, ...) are needed.

@@

\subsection{Impact of Generalized Loop-invariant Code Motion}

\newcommand{\massmatrixtwodresultsnorms}{
\begin{tabularx}{1.0\textwidth}{cccccc|cccc|cccc|cccc}
\cline{3-18}
& & \multicolumn{8}{c}{\texttt{Sandy Bridge}} & \multicolumn{8}{c}{\texttt{Xeon Phi}}\\
\cline{3-18}
& & \multicolumn{4}{c}{\texttt{licm}} & \multicolumn{4}{c}{\texttt{licm-ap}} & \multicolumn{4}{c}{\texttt{licm}} & \multicolumn{4}{c}{\texttt{licm-ap}} \\
\cline{1-18}
& & \texttt{p1} & \texttt{p2} & \texttt{p3} & \texttt{p4} & \texttt{p1} & \texttt{p2} & \texttt{p3} & \texttt{p4} & \texttt{p1} & \texttt{p2} & \texttt{p3} & \texttt{p4} & \texttt{p1} & \texttt{p2} & \texttt{p3} & \texttt{p4} \\[0.1cm]
\texttt{Helmholtz} & \texttt{triangle} & 1.05 & 1.46 & 1.68 & 1.67 & 1.32 & 1.88 & 2.87 & 4.13 & 1,49 & 1,06 & 1,05 & 1,17 & 1,50 & 2,41 & 1,30 & 1,96 \\
\texttt{Helmholtz} & \texttt{tetrahedron} & 1.36 & 2.10 & 2.64 & 2.27 & 1.35 & 3.32 & 2.66 & 3.27 & 1,28 & 1,29 & 2,05 & 1,73 & 1,41 & 1,50 & 2,79 & 2,81 \\
\texttt{Helmholtz} & \texttt{prism} & 2.16 & 2.28 & 2.45 & 2.06 & 2.63 & 2.74 & 2.43 & 2.75 & 1,04 & 2,26 & 1,93 & 1,64 & 2,38 & 2,47 & 2,15 & 1,71\\[0.1cm]
%\cline{1-18}
\texttt{Diffusion} & \texttt{triangle} & 1.09 & 1.68 & 1.97 & 1.64 & 1.38 & 1.99 & 3.07 & 4.28 & 1,07 & 1,06 & 1,18 & 1,16 & 1,08 & 1,88 & 1,20 & 1,97\\
\texttt{Diffusion} & \texttt{tetrahedron} & 1.30 & 2.20 & 3.12 & 2.60 & 1.41 & 3.70 & 3.18 & 3.82 & 1,00 & 1,38 & 2,02 & 1,74 & 1,05 & 1,51 & 2,76 & 3,00\\
\texttt{Diffusion} & \texttt{prism} & 2.15 & 1.82 & 2.71 & 2.32 & 2.55 & 3.13 & 2.73 & 2.69 & 1,11 & 2,16 & 1,85 & 2,83 & 2,41 & 2,52 & 2,05 & 2,48\\[0.1cm]
%\cline{1-18}
\texttt{Burgers} & \texttt{triangle} & 1.53 & 1.81 & 2.68 & 2.46 & 1.56 & 2.28 & 2.61 & 2.77 & 1,21 & 1,42 & 2,34 & 2,97 & 2,84 & 2,26 & 3,96 & 4,27 \\
\texttt{Burgers} & \texttt{tetrahedron} & 1.61 & 2.24 & 1.69 & 1.59 & 1.61 & 2.10 & 1.60 & 1.78 & 1,01 & 2,55 & 0,98 & 1,21 & 1,48 & 3,83 & 1,55 & 1,29 \\
\texttt{Burgers} & \texttt{prism} & 2.11 & 2.20 & 1.66 & 1.32 & 2.19 & 2.32 & 1.64 & 1.42 & 1,39 & 1,56 & 1,18 & 1,04 & 2,18 & 2,82 & 1,24 & 1,25 \\
\cline{1-18}
\end{tabularx}
}

@@

Table~\ref{table:perf-results-licm} illustrates the performance improvement obtained on the Sandy Bridge and Xeon Phi machines when \emph{licm}, data alignment, and padding are used. We distinguish between \emph{licm} and \emph{licm-ap}; the latter makes use of padding and data alignment. Inspection of assembly code generated by \texttt{icc} confirmed all limitations described in Section~\ref{sec:code-transf-licm}: only sub-expressions invariant to the inner loop are hoisted and their execution is not vectorized. Padding and data alignment enhance, in general, the quality of auto-vectorization. Sometimes the run-time of \emph{licm-ap} is similar to that of \emph{licm} because the element matrix size is already a multiple of the vector length, so no scalar reminder loop has to be introduced. Occasionally \emph{licm-ap} is slower than \emph{licm} (e.g. in Burgers $p=3$ on the Sandy Bridge). One possible explanation is that the large number of aligned temporaries introduced by \emph{licm} induce cache associativity conflicts.

@@

\section{Related Work}
\label{sec:related-work}
The finite element method is extensively used to approximate solutions of PDEs. Well-known frameworks and applications include nek5000~\cite{nek5000-web-page}, the FEniCS project~\cite{Fenics}, Fluidity~\cite{fluidity_manual_v4}, and of course Firedrake. Numerical integration is usually employed to implement the local assembly phase. The recent introduction of DSLs to decouple the finite element specification from its underlying implementation facilitated, however, the development of novel approaches. Methods based on tensor contraction~\cite{FFC-Compiler} and symbolic manipulation~\cite{Francis} have been developed. We have designed COFFEE to specifically target numerical integration because it has been demonstrated that it remains the optimal choice for a wide class of problems~\cite{quadrature1}.
